---
title: "Estimating delays under recall-biased symptom onset reporting"
---

## Setup
```{r}
knitr::opts_chunk$set(cache = FALSE)
library(dplyr)
library(ggplot2)
source(here::here("R", "mess-up-linelist.R"))
source(here::here("R", "messy-linelist.R"))

library(primarycensored)
library(cmdstanr)
```

## Primarycensored setup
```{r}
pcd_model <- pcd_cmdstan_model(cpp_options = list(stan_threads = FALSE))
```
```{r}
pcd_data_function <- function(delay_data) {
  pcd_data <- pcd_as_stan_data(
    delay_data,
    delay = "observed_delay",
    delay_upper = "observed_delay_upper",
    relative_obs_time = "D",
    pwindow = "pwindow",
    dist_id = pcd_stan_dist_id("lognormal", "delay"),
    primary_id = pcd_stan_dist_id("uniform", "primary"),
    param_bounds = list(lower = c(-Inf, 0), upper = c(Inf, Inf)),
    primary_param_bounds = list(lower = numeric(0), upper = numeric(0)),
    priors = list(location = c(1, 0.5), scale = c(1, 0.5)),
    primary_priors = list(location = numeric(0), scale = numeric(0)),
    use_reduce_sum = FALSE # use within chain parallelisation
  )
}
```

## Data preparation
```{r}
ll_clean <- readr::read_csv(here::here("data-raw", "clean_linelist.csv")) |>
  filter(asymptomatic == FALSE, !is.na(date_onset), !is.na(date_admission))

day_90_observed <- 1
day_50_observed <- 7
day_10_observed <- 2 * day_50_observed - day_90_observed

ll_recall <- add_recall_missing(
  ll_clean,
  day_90_observed = day_90_observed,
  day_50_observed = day_50_observed,
  rng_seed = 2026
)

ll <- new_messy_linelist(ll_clean)
ll[, date_onset := ll_recall$date_onset]
```

## The recall bias mechanism
```{r}
recall_curve <- data.frame(
  true_delay = 0:max(as.numeric(ll$date_admission_true - ll$date_onset_true), na.rm = TRUE)
) |>
  mutate(
    p_observed_recall = plogis(
      qlogis(0.9) +
        (true_delay - day_90_observed) *
        (qlogis(0.5) - qlogis(0.9)) /
        (day_50_observed - day_90_observed)
    )
  )

recall_empirical <- ll |>
  transmute(
    true_delay = as.numeric(date_admission_true - date_onset_true),
    onset_observed = !is.na(date_onset)
  ) |>
  filter(true_delay >= 0) |>
  group_by(true_delay) |>
  summarise(
    p_observed_empirical = mean(onset_observed),
    n = n(),
    .groups = "drop"
  )

ggplot() +
  geom_line(
    data = recall_curve,
    aes(x = true_delay, y = p_observed_recall),
    linewidth = 1.1,
    color = "#125c87"
  ) +
  geom_point(
    data = recall_empirical,
    aes(x = true_delay, y = p_observed_empirical, size = n),
    color = "#a00000",
    alpha = 0.7
  ) +
  geom_vline(xintercept = c(day_90_observed, day_50_observed, day_10_observed),
             linetype = c("dashed", "dashed", "dotted"),
             color = c("#125c87", "#800000", "grey40")) +
  theme_bw() +
  labs(
    x = "True onset-to-admission delay (days)",
    y = "Probability onset date is observed",
    size = "Records"
  )
```

## Comparing approaches to recall bias

### Unsuitable approach 1: Complete-case analysis
```{r}
delay_data <- ll |>
  filter(!is.na(date_onset)) |>
  mutate(delay = as.numeric(date_admission - date_onset)) |>
  filter(delay >= 0) |>
  group_by(delay) |>
  summarise(n = n()) |>
  ungroup() |>
  transmute(
    observed_delay = delay,
    observed_delay_upper = delay + 1,
    pwindow = 1,
    D = max(delay, na.rm = TRUE) + 2,
    n = n
  )

pcd_fit_complete_case <- pcd_model$sample(
  data = pcd_data_function(delay_data),
  chains = 4,
  parallel_chains = 2,
  threads_per_chain = 2,
  refresh = ifelse(interactive(), 50, 0),
  show_messages = interactive()
)
```

### Unsuitable approach 2: Ad-hoc imputation using date of report
```{r}
ll_impute_report <- data.table::copy(ll)
do_impute_from_other(ll_impute_report, "date_onset", "date_reporting")

delay_data <- ll_impute_report |>
  mutate(delay = as.numeric(date_admission - date_onset)) |>
  filter(delay >= 0) |>
  group_by(delay) |>
  summarise(n = n()) |>
  ungroup() |>
  transmute(
    observed_delay = delay,
    observed_delay_upper = delay + 1,
    pwindow = 1,
    D = max(delay, na.rm = TRUE) + 2,
    n = n
  )

pcd_fit_impute_report <- pcd_model$sample(
  data = pcd_data_function(delay_data),
  chains = 4,
  parallel_chains = 2,
  threads_per_chain = 2,
  refresh = ifelse(interactive(), 50, 0),
  show_messages = interactive()
)
```

### Suitable approach 1: Broad censoring model
Assume missing onset dates are only known to have happened within 8 weeks before
admission.
```{r}
delay_data <- ll |>
  mutate(
    earliest_onset_date = date_admission - day_10_observed,
    latest_onset_date = date_admission
  ) |>
  mutate(
    delay = as.numeric(date_admission - if_else(is.na(date_onset), earliest_onset_date, date_onset)),
    pwindow = if_else(is.na(date_onset), as.numeric(latest_onset_date - earliest_onset_date) + 1, 1)
  ) |>
  filter(delay >= 0) |>
  group_by(delay, pwindow) |>
  summarise(n = n()) |>
  ungroup() |>
  transmute(
    observed_delay = delay,
    observed_delay_upper = delay + 1,
    pwindow = pwindow,
    D = max(delay, na.rm = TRUE) + 2,
    n = n
  )

pcd_fit_censor_broad <- pcd_model$sample(
  data = pcd_data_function(delay_data),
  chains = 4,
  parallel_chains = 2,
  threads_per_chain = 2,
  refresh = ifelse(interactive(), 50, 0),
  show_messages = interactive()
)
```

### Suitable approach 2: Recall-informed sensitivity analysis
Use the assumed recall curve to define a tighter onset window. Under the
logistic recall model, the day at which only 10% of dates are still observed is
`2 * day_50_observed - day_90_observed`.
```{r}
delay_data <- ll |>
  mutate(
    earliest_onset_date = date_admission - day_10_observed,
    latest_onset_date = date_admission
  ) |>
  mutate(
    delay = as.numeric(date_admission - if_else(is.na(date_onset), earliest_onset_date, date_onset)),
    pwindow = if_else(is.na(date_onset), as.numeric(latest_onset_date - earliest_onset_date) + 1, 1)
  ) |>
  filter(delay >= 0) |>
  group_by(delay, pwindow) |>
  summarise(n = n()) |>
  ungroup() |>
  transmute(
    observed_delay = delay,
    observed_delay_upper = delay + 1,
    pwindow = pwindow,
    D = max(delay, na.rm = TRUE) + 2,
    n = n
  )

pcd_fit_censor_recall <- pcd_model$sample(
  data = pcd_data_function(delay_data),
  chains = 4,
  parallel_chains = 2,
  threads_per_chain = 2,
  refresh = ifelse(interactive(), 50, 0),
  show_messages = interactive()
)
```

## Results
```{r}
pcd_fit_truth <- pcd_model$sample(
  data = ll_clean |>
    mutate(delay = as.numeric(date_admission - date_onset)) |>
    filter(delay >= 0) |>
    group_by(delay) |>
    summarise(n = n()) |>
    ungroup() |>
    transmute(
      observed_delay = delay,
      observed_delay_upper = delay + 1,
      pwindow = 1,
      D = max(delay, na.rm = TRUE) + 2,
      n = n
    ) |>
    pcd_data_function(),
  chains = 4,
  parallel_chains = 2,
  threads_per_chain = 2,
  refresh = ifelse(interactive(), 50, 0),
  show_messages = interactive()
)
```

```{r}
rvars <- lapply(
  list(
    pcd_fit_complete_case,
    pcd_fit_impute_report,
    pcd_fit_censor_broad,
    pcd_fit_censor_recall
  ),
  function(x) posterior::as_draws_rvars(x$draws("params"))
)

truth_mean_delay <- pcd_fit_truth$summary(c("params[1]", "params[2]"))
truth_mean_delay <- exp(truth_mean_delay$mean[1] + truth_mean_delay$mean[2]^2 / 2)

results_df <- data.frame(
  Estimate = forcats::fct_inorder(c(
    "Complete cases\n(unsuitable)",
    "Report-date imputation\n(unsuitable)",
    "Broad censoring window\n(suitable)",
    "Recall-informed bound\n(suitable)"
  )),
  mean_delay = t(do.call(cbind, lapply(
    rvars,
    function(x) exp(x$params[[1]] + x$params[[2]]^2 / 2)
  )))
)

ggplot(results_df) +
  ggdist::stat_halfeye(aes(fill = Estimate, color = Estimate, xdist = mean_delay), alpha = 0.6) +
  geom_vline(xintercept = truth_mean_delay, color = "black", linetype = "dashed") +
  scale_color_manual(values = c("#800000", "orange", "#125c87", "#2A9D8F")) +
  scale_fill_manual(values = c("#a00000", "orange", "#1a80bb", "#2A9D8F")) +
  theme_bw() +
  theme(legend.position = "top") +
  labs(x = "Mean delay (days)", y = "Density")

# ggsave(here::here("vignettes", "figures", "issue_recall_bias_approaches.png"), width = 10, height = 6)
```
