---
title: "Estimating delays from line list data with imputed symptom onset date"
---

## Setup
```{r}
knitr::opts_chunk$set(cache = TRUE)
library(dplyr)
library(ggplot2)
source(here::here("R", "messy-linelist.R"))

library(primarycensored)
library(cmdstanr)
```

## Primarycensored setup
```{r}
pcd_model <- pcd_cmdstan_model(cpp_options = list(stan_threads = FALSE))
```
```{r}
pcd_data_function <- function(delay_data) {
  pcd_data <- pcd_as_stan_data(
    delay_data,
    delay = "observed_delay",
    delay_upper = "observed_delay_upper",
    relative_obs_time = "D",
    pwindow = "pwindow",
    dist_id = pcd_stan_dist_id("lognormal", "delay"),
    primary_id = pcd_stan_dist_id("uniform", "primary"),
    param_bounds = list(lower = c(-Inf, 0), upper = c(Inf, Inf)),
    primary_param_bounds = list(lower = numeric(0), upper = numeric(0)),
    priors = list(location = c(1, 0.5), scale = c(1, 0.5)),
    primary_priors = list(location = numeric(0), scale = numeric(0)),
    use_reduce_sum = FALSE # use within chain parallelisation
  )
}
```

## Data preparation
```{r}
ll_clean <- readr::read_csv(here::here("data-raw", "clean_linelist.csv"))
```

Setup a messy linelist
```{r}
ll <- new_messy_linelist(ll_clean)
be_missing(ll, "date_onset", prob = 0.5) # date of symptom onset missing with 50% probability
```

## Estimating the onset to admission delay distribution

### Messy analysis: Imputed using the date of report
```{r}
ll_messy <- copy(ll)
do_impute_from_other(ll_messy, "date_onset", "date_reporting")

delay_data <- ll_messy |> 
  mutate(delay = as.numeric(date_admission - date_onset)) |>
  filter(delay >= 0) |>
  group_by(delay) |>
  summarise(n = n()) |>
  ungroup() |>
  transmute(
    observed_delay = delay,
    observed_delay_upper = delay + 1,
    pwindow = 1,
    D = max(delay, na.rm = TRUE) + 2,
    n = n
    )

pcd_fit_messy <- pcd_model$sample(
  data = pcd_data_function(delay_data),
  chains = 4,
  parallel_chains = 2, # Run 2 chains in parallel
  threads_per_chain = 2, # Use 2 cores per chain
  refresh = ifelse(interactive(), 50, 0),
  show_messages = interactive()
)
```

### Messy analysis 2: Only symptomatic, but still imputed using the date of report
```{r}
ll_messy2 <- copy(ll)
do_impute_from_other(ll_messy2, "date_onset", "date_reporting")

delay_data <- ll_messy2 |> 
  filter(asymptomatic == FALSE) |> 
  mutate(delay = as.numeric(date_admission - date_onset)) |>
  filter(delay >= 0) |>
  group_by(delay) |>
  summarise(n = n()) |>
  ungroup() |>
  transmute(
    observed_delay = delay,
    observed_delay_upper = delay + 1,
    pwindow = 1,
    D = max(delay, na.rm = TRUE) + 2,
    n = n
    )

pcd_fit_messy2 <- pcd_model$sample(
  data = pcd_data_function(delay_data),
  chains = 4,
  parallel_chains = 2, # Run 2 chains in parallel
  threads_per_chain = 2, # Use 2 cores per chain
  refresh = ifelse(interactive(), 50, 0),
  show_messages = interactive()
)
```

### Correct analysis
do not impute, instead assume:
- earliest onset: earliest onset date in dataset, but not earlier than 8 weeks before admission
- latest onset: maximum of date of admission / date of report
- date of admission: known (1 day censoring)
```{r}
delay_data <- ll |> 
  filter(asymptomatic == FALSE) |>
  mutate(
    earliest_onset_date = pmax(min(date_onset, na.rm = TRUE), date_admission - 7*4*2),
    latest_onset_date = pmax(date_admission, date_reporting, na.rm = TRUE)
  ) |> 
  mutate(delay = as.numeric(date_admission - ifelse(is.na(date_onset), earliest_onset_date, date_onset))) |>
  mutate(pwindow = ifelse(is.na(date_onset), as.numeric(latest_onset_date - earliest_onset_date) + 1, 1)) |> 
  filter(delay >= 0) |>
  group_by(delay, pwindow) |>
  summarise(n = n()) |>
  ungroup() |>
  transmute(
    observed_delay = delay,
    observed_delay_upper = delay + 1,
    pwindow = pwindow,
    D = max(delay, na.rm = TRUE) + 2,
    n = n
    )

pcd_fit_correct <- pcd_model$sample(
  data = pcd_data_function(delay_data),
  chains = 4,
  parallel_chains = 2, # Run 2 chains in parallel
  threads_per_chain = 2, # Use 2 cores per chain
  refresh = ifelse(interactive(), 50, 0),
  show_messages = interactive()
)
```

## Results
```{r}
rvars <- lapply(list(pcd_fit_messy, pcd_fit_messy2, pcd_fit_correct), function(x) posterior::as_draws_rvars(x$draws("params")))

results_df <- data.frame(
  Estimate = forcats::fct_inorder(c("Ad-hoc imputation\n(with asymptomatic)", "Ad-hoc imputation\n(only symptomatic)", "Missingness model\n(only symptomatic)")),
  mean_delay = t(do.call(cbind, lapply(rvars, function(x) exp(x$params[[1]] + x$params[[2]]^2 / 2))))
)

ggplot(results_df) +
  ggdist::stat_halfeye(aes(fill = Estimate, color = Estimate, xdist = mean_delay), alpha = 0.6) +
  geom_vline(xintercept = exp(1.5 + 0.5^2 / 2), color = "black", linetype = "dashed") +
  scale_color_manual(values = c("#800000", "orange", "#125c87")) +
  scale_fill_manual(values = c("#a00000", "orange", "#1a80bb")) +
  theme_bw() +
  theme(legend.position = "top") +
  labs(x = "Mean delay (days)", y = "Density")

ggsave(here::here("vignettes", "figures", "issue_imputation.png"), width = 10, height = 6)
```
